---
title: 'Prosper Loan Data Analysis'
author: 'Jörg Horchler'
output:
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
    highlight: pygments
  html_document:
    df_print: paged
    highlight: pygments
    number_sections: true
    theme: flatly
    toc: true
    toc_float: true
    dev: quartz_png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

```{r packages, message=FALSE, warning=FALSE}
# Load libraries used for plotting
library(ggplot2)
library(pander)
library(ggthemes)
library(scales)
library(ggcorrplot)
library(ggalt)
# wrangling/grouping etc.
library(dplyr)
library(forcats)
library(lubridate)
# nicer output
library(knitr)
library(summarytools)
```

```{r set_options, message=FALSE, warning=FALSE}
# theme defaults
panderOptions('graph.grid.color', 'red')
panderOptions('graph.grid.lty', 'dotted')
# geom colors
update_geom_defaults('bar', list(colour = 'black', fill = '#f0f8ff'))
update_geom_defaults('area', list(colour = 'black', fill = '#89cff0'))
update_geom_defaults('boxplot', list(colour = 'black', fill = '#bcd4e6'))
update_geom_defaults('point', list(colour = '#3d2b1f'))
```

***

# Introduction

This document explores a dataset about loans which was collected from
[Prosper](https://www.prosper.com). 
The [datafile](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/prosperLoanData.csv&sa=D&ust=1537305070260000) was last updated on 03/11/2014.

Prosper is an online marketplace where creditworthy borrowers can request a loan
and investors can invest in notes (or pieces) of each loan. Investors receive a
portion of those payments directly to their Prosper account proportional to
their pro rata share of the loan.

The variables in this dataset include loan, including loan amount, borrower rate
(or interest rate), current loan status, borrower income, borrower employment 
status, borrower credit history, and the latest payment information.

The objective of this report is to examine how the interest rate is calculated.
This can be described as what variables are used to examine the risk factors of
a borrower.

***

# Dataset description

```{r load_data}
loan_data <- read.csv('prosperLoanData.csv', na.strings = '')
```

## Structure

The structure is:

```{r show_structure}
str(loan_data)
```

This output already shows some points to note: 

1. The date fields like `ListingCreationDate`, `ClosedDate` or 
   `DateCreditPulled` were imported as `Factor`. These fields should be 
   converted to dates.
2. `CreditGrade` should be sorted from low risk to high risk. This order feels
   more naturally.
3. Two other fields hold `Ratings` as well. These are 
   `ProsperRating (numeric)` which encodes `ProsperRating (Alpha)`. These should
   be sorted as well.
4. The same should be done for `IncomeRange` and `ProsperScore`.
5. Some variables hold `True` or `False`. For example `IncomeVerifiable` and
   `IsBorrowerHomeowner`. These should be converted to bools.
6. There are a lot `NA` values. How to handle these will be decided during the
   analysis.

These tasks will be done during the analysis if needed. 

## Dimension

In the above output it is visible that the dataset contains 113,937 observations 
of 81 variables.

***

# Getting the feet wet

In this section the report contains an exploration of various 
but not all variables in the dataset. The goal is to explore what variables 
might be important for the examination of the risk rating system.

## Background Information

The history of Prosper and the obligation to register with SEC can be read at 
[Harvard Business Law Review][HBLR]. To sum up, when Prosper started its
business it started in a unregulated market. P2P lending was new and fraught 
with risk to lenders, who were largely individuals rather than traditional
institutional creditors.

From 2006 to 2008 Prosper was charging off more than 20% of loans issued. On
November 2008 Securities and Exchange Commission (SEC) entered a 
cease-and-desist order against Prosper. From then on Prosper had to be 
registered with the SEC to comply with federal securities laws.

Finally in July 2009, Prosper reopened their website after having obtained SEC 
registration. Starting with that Prosper rolled out its new proprietary credit
rating system what is called **Prosper Rating**. It is based on two scores:

* The credit score from a credit reporting agency.
* The second is a Prosper Score of which is based upon a custom risk model using
  Prosper data to predict the probability of a loan going *"bad"*.

## Timeframe

The first important variable is what timeframe is included in the dataset. As
written above Prosper relaunched in 2009. As `ProsperRating` is only used with
data after this relaunch it is important to use only data after the relaunch.

For that the *Factor* `ListingCreationDate` is converted as date and is then 
plotted with the number of listings created by day.

```{r convert_dates}
loan_data$ListingCreationDate <- as.Date(loan_data$ListingCreationDate,
                                         format = '%Y-%m-%d %H:%M:%S')
```

```{r listing_creation_date_by_day}
ggplot(loan_data, aes(ListingCreationDate)) +
  geom_histogram(binwidth = 1) +
  theme_pander()
```

Two things are visible here: 

1. A significant drop of listings created in 2009 - the year of the SEC
   registration.
2. One large peak at the beginning of 2011.

The first point is explained above.

```{r create_post_sec_data_frame}
post_sec_relaunch <- loan_data %>%
  filter(ListingCreationDate >= as.Date('2009-07-29'))
```

The large peak at the beginning of 2011 is at date:

```{r peak_2011_beginning}
post_sec_relaunch %>%
  filter(year(ListingCreationDate) == '2011') %>%
  count(ListingCreationDate) %>%
  top_n(1, n) %>%
  select('ListingCreationDate') %>%
  pull()
```

Research did not unveil what happened on this day and why there was that
significant raise in listings created.

```{r listing_creation_date_range_post_sec_relaunch, results = 'hide'}
summary(post_sec_relaunch$ListingCreationDate)
```

After the relaunch the range of `ListingCreationDate` is from **2009-07-29** to
**2014-03-10**.

Let's see the distribution by month in this period:

```{r post_sec_relaunch_monthly_distribution}
ggplot(post_sec_relaunch, aes(ListingCreationDate)) +
  geom_histogram(binwidth = 30) +
  theme_pander()
```

Obviously P2P lending has gotten more popular in 2013.

I'd like to see later on how various variables changed over time.

## BorrowerRate

As next information in this section the `Interest Rate` that borrowers is 
offered is examined. This is stored as `BorrowerRate` and is the most important 
variable for borrowers *and* lenders. This is because lender yield is
`BorrowerRate - 0.01`.

```{r borrower_rate_summary, results = 'asis'}
descr(post_sec_relaunch$BorrowerRate, style = 'rmarkdown', 
      omit.headings = TRUE)
```

Prosper advertizes its low interest rate. In this dataset the range is from
**0.04** to **0.36**. The average rate is about **0.20**.

That's quite high but is just the average across the whole dataset. On the other
hand **4%** is very low compared to credit card interest rates as seen on
[Federal Reserve Economic Data][TERMCBCCALLNS] of the timeframe covered in this
dataset. According to that the Interest Rate on Credit Card Plans of
Commercial Banks was between **14.5%** to around **12%**.

```{r borrower_rate_distribution}
ggplot(post_sec_relaunch, aes(BorrowerRate)) +
  geom_histogram(binwidth = 0.01) +
  theme_pander() +
  scale_x_continuous(label = percent)
```

As we can see the distribution is multimodal. The majority of the loans do have 
a rate of around **32%** what is very high. The top 5 interest rates are:

```{r top_5_worse_borrower_ratings}
post_sec_relaunch %>%
  count(BorrowerRate) %>%
  top_n(., 5, n) %>%
  arrange(desc(n)) %>%
  kable(col.names = c('Borrower Rate','Listings'))
```

A higher `BorrowerRate` means a worse rating.

```{r top_5_worse_borrower_ratings_var}
top_5_worse_borrower_ratings <- post_sec_relaunch %>%
  count(BorrowerRate) %>%
  top_n(., 5, n) %>%
  select(BorrowerRate) %>%
  pull()
```

```{r prosper_rating_of_top_5_worse_borrower_ratings}
# reorder the factor
rate_order <- c('AA', 'A', 'B', 'C', 'D', 'E', 'HR')
post_sec_relaunch$ProsperRating..Alpha. <- 
  factor(post_sec_relaunch$ProsperRating..Alpha., rate_order)
post_sec_relaunch %>%
  filter(BorrowerRate %in% top_5_worse_borrower_ratings) %>%
  count(ProsperRating..Alpha.) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(., aes(ProsperRating..Alpha., n)) +
  geom_bar(stat = 'identity') + 
  geom_text(aes(y = n, label = percent(proportion)), stat = 'identity',
            vjust = -0.5) +
  theme_pander()
```

One can see a glimpse on the relationship between `ProsperRating` and
`BorrowerRate`. As seen here most *bad* interest rates (`BorrowerRate`) do have
a *bad* `ProsperRating` as well.

I am left wondering how `BorrowerRate` is related to other variables like
`ProsperScore` or `CreditRange`.

## ProsperScore

`ProsperScore` is defined as *a custom risk score built using historical Prosper
data*. It is used to calculate `ProsperRating`.

```{r prosper_score_summary, results = 'asis'}
descr(post_sec_relaunch$ProsperScore, style = 'rmarkdown', 
      omit.headings = TRUE)
```

`ProsperScore` ranges from **1** to **11**, with 11 being the best, or lowest
risk score. The median in this dataset is **6**. The **mean** doesn't make 
sense here because Prosper doesn't use fractions of scores. Let's see whether
this variable follows a normal distribution:

```{r prosper_score_dist}
ggplot(post_sec_relaunch, aes(ProsperScore)) +
  geom_histogram(binwidth = 1) +
  geom_text(aes(y = ..count.., label = percent(..count.. / sum(..count..))),
            stat = 'count', vjust = -0.5) +
  theme_pander() +
  scale_x_continuous(breaks = seq(1, 11, 1))
```

This distribution is multimodal having the maximum at a score of **4** followed
by **6** and **8**. But it is visible that the majority of loans is spread
around the mean. 

Again I wonder how this score influences `BorrowerRate`.

## ProsperRating

`ProsperRating` is another variable used to calculate the `BorrowerRate`
customers is offered. It is available in two variants: 

* `ProsperRating (Alpha)` is encoded by
* `ProsperRating (numberic)`

What are the summary statistics and visualization of `ProsperRating (numeric)`?

```{r prosper_rating_summary, results = 'asis'}
descr(post_sec_relaunch$ProsperRating..numeric., style = 'rmarkdown', 
      omit.headings = TRUE)
```

As seen the rating is from **1** to **7** what is the coding for *HR* to *AA*.
That means that **1** is the worst and **7** is the best rating. The median
rating is **4** what encodes a **C** rating. Again as no fractions of the rating
exist the mean doesn't make sense.

```{r prosper_rating_dist}
ggplot(post_sec_relaunch, aes(ProsperRating..numeric.)) +
  geom_bar() +
  geom_text(aes(y = ..count.., label = percent(..count.. / sum(..count..))), 
            stat = 'count', vjust = -0.5) +
  scale_x_continuous(breaks = seq(1, 7, 1)) +
  theme_pander()
```

This distribution is normal distributed what I'd expected here because I think 
that most borrowers at Prosper are having an average income and 
creditworthiness.

I wonder how that relates to the prosper score and the borrower rate. These
variables do not follow a normal distribution.

## CreditScoreRangeLower and CreditScoreRangeUpper

I suspect the credit score Prospers pulls from a agency influences the
interest rate. Let's first see again the descriptive statistics.

```{r credit_score_range_lower_summary, results = 'asis'}
descr(post_sec_relaunch$CreditScoreRangeLower, style = 'rmarkdown', 
      omit.headings = TRUE)
```

```{r credit_score_range_upper_summary, results = 'asis'}
descr(post_sec_relaunch$CreditScoreRangeUpper, style = 'rmarkdown', 
      omit.headings = TRUE)
```

As seen `CreditScoreRangeUpper` is just **19** points higher than 
`CreditScoreRangeLower`. Both have a standard deviation of **47.07**, the *IQR*
is **60** and so on. *Min* and *Max* are from **600**/**619** to 
**880**/**899**, respectively. 

Let's see just the distribution of the lower range:

```{r credit_score_lower_dist}
ggplot(post_sec_relaunch, aes(CreditScoreRangeLower)) +
  geom_histogram(binwidth = 19) +
  scale_x_continuous(breaks = seq(600, 900, 15)) +
  theme_pander()
```

This variable is right skewed having the modal at **660**. As seen above there
are no **NA** values and not outliers. We see here as well (and from the
statistics above) that Prosper only allows borrows by people having a 
credit score above 600. This is one of the consequences described above.

I suspect the strongest correlation between `BorrowerRate` and 
`CreditScoreRangeLower`.

## StatedMonthlyIncome

`StatedMonthlyIncome` is a allegation of the borrower about his income.

```{r stated_monthly_income_summary, results = 'asis'}
descr(post_sec_relaunch$StatedMonthlyIncome, style = 'rmarkdown', 
      omit.headings = TRUE)
```

The statistics show a range from **$0** to **$1,750,002.92**. 

***

> Wow!

***

The average stated income is **$5,932.96** and there are no **NA** values.
**Q3** is **$7083.33**. As I suspect that **$1,750,002.92** is an outlier let's
count the listings above Q3.

```{r stated_monthly_income_above_q3}
post_sec_relaunch %>%
  filter(StatedMonthlyIncome > quantile(StatedMonthlyIncome, 0.75)) %>%
  nrow()
```

That's still a lot of rows. So what are the top 10 monthly incomes?

```{r top_10_monthly_incomes_table}
post_sec_relaunch %>%
  filter(StatedMonthlyIncome > quantile(StatedMonthlyIncome, 0.75)) %>%
  select(StatedMonthlyIncome) %>%
  top_n(., 10, StatedMonthlyIncome) %>%
  arrange(desc(StatedMonthlyIncome)) %>%
  kable()
```

Now let's see this one row:

```{r maximum_stated_income_row}
# show this one row as dataframe
post_sec_relaunch %>%
  filter(StatedMonthlyIncome == max(StatedMonthlyIncome))
```

The listing was created by someone being a **Self-employed** **Professional**
who is a Home Owner and who's income is not verifiable. The 
**LoanOriginalAmount** is about **4000** Dollars. The rating is at 
**High Risk**. From these information it is not clear whether
this a an outlier. But this has to be noted because such values skew the
distribution.

That is the distribution without this row and without incomes of **0**:

```{r stated_income_histogram_subset}
post_sec_relaunch %>%
  filter(StatedMonthlyIncome > 0 &
         StatedMonthlyIncome < max(StatedMonthlyIncome)) %>%
  ggplot(., aes(StatedMonthlyIncome)) + 
  geom_histogram(binwidth = 10000) +
  scale_x_continuous(label = dollar, breaks = seq(0, 618548, 75000)) +
  theme_pander()
```

As this plot does not reveal that much let's exclude more rows:

```{r stated_income_histogram_subset_below_75000}
post_sec_relaunch %>%
  filter(StatedMonthlyIncome < 20000) %>%
  ggplot(., aes(StatedMonthlyIncome)) + 
  geom_histogram(binwidth = 1000, boundary = 0) +
  scale_x_continuous(label = dollar, breaks = seq(0, 20000, 1000)) +
  theme_pander() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

This is an expected right skewed plot. The *binwidth* here is **$1000**.

I wonder about the ratings and the scores of the members having more than
**$20,000** stated monthly income.

## PublicRecordsLast12Months

`PublicRecordsLast12Months` is one of the variables that should not occur in 
this dataset as people are not allowed to join Prosper having a Public Record
within the last 12 months. But perhaps there are some loans of people joining
Prosper having no public record but got one being a member.

```{r public_records_last_12_months_summary, results = 'asis'}
descr(post_sec_relaunch$PublicRecordsLast12Months, style = 'rmarkdown', 
      omit.headings = TRUE)
```

As expected the minimum is **0**. But the maximum is **20**. Is it possible to
get that much public records?

```{r public_records_last_12_months_distribution}
ggplot(post_sec_relaunch, aes(as.factor(PublicRecordsLast12Months))) +
  geom_bar() +
  scale_y_continuous(breaks = seq(0, 90000, 10000)) +
  geom_text(aes(y = ..count.., label = ..count..),
            stat = 'count', vjust = -0.4) +
  theme_pander()
```

There are not that much listings having a public record. Even **621** listings
having **1** public record is not that much.

What is the `LoanStatus` of all of these listings?

```{r public_record_loan_status}
# first sort LoanStatus
post_sec_relaunch$LoanStatus <- factor(post_sec_relaunch$LoanStatus, 
                                       levels=c('Completed',
                                                'FinalPaymentInProgress',
                                                'Current',
                                                'Past Due (1-15 days)',
                                                'Past Due (16-30 days)',
                                                'Past Due (31-60 days)',
                                                'Past Due (61-90 days)',
                                                'Past Due (91-120 days)',
                                                'Past Due (>120 days)',
                                                'Defaulted',
                                                'Chargedoff',
                                                'Cancelled'))
# show frequency
post_sec_relaunch %>%
  filter(PublicRecordsLast12Months > 0) %>%
  count(LoanStatus) %>%
  kable(col.names = c('LoanStatus', 'Frequency'))
```

Having the majority of loans without public records I do not expect that this
variable is correlated to `BorrowerRate`. But this is a point to examine later
on.

## InquiriesLast6Months

`InquiriesLast6Months` is the number of inquiries in the last 6 months at the 
time the credit profile was pulled by Prosper. This is of interest to see how 
many credit bureau inquiries were requested and in addition is one variable 
Prospers states as [criteria][MinimumCriteria] to borrow on Prosper. This is 
defined as *"Fewer than five credit bureau inquiries within the last 6 months"*.

Due to that there should be no listing having more than 5 inquiries.

```{r inquiries_summary, results = 'asis'}
descr(post_sec_relaunch$InquiriesLast6Months, style = 'rmarkdown', 
      omit.headings = TRUE)
```

The range of this variable is from **0** to **27** and there are no **NA**
values. As **Q3** is at **1** most of the values in this variable must be **0**.

A histogram showing that is:

```{r inquiries_histogram}
ggplot(post_sec_relaunch, aes(InquiriesLast6Months)) +
  geom_histogram(binwidth = 1) +
  theme_pander() +
  scale_x_continuous(breaks = seq(0, 27, 1))
```

The following table shows the proportions:

```{r inquiries_proportion}
# choosed to display a table because a pie chart looked not that clear
post_sec_relaunch %>%
  count(InquiriesLast6Months) %>%
  mutate(freq = round(100 * (n / sum(n)), 2)) %>%
  kable(col.names = c('Inquiries Last 6 Months', 'Frequency', 'Proportion'))
```

Roughly **50%** of the borrowers in the dataset do have **0** inquiries in the
last six months. Overall **97%** of the dataset falls below the minimum
requirement of having fewer than five inquiries in the last six months.

What is the distribution above the minimum requirement:

```{r inquiries_histogram_out_of_range}
post_sec_relaunch %>%
  filter(InquiriesLast6Months > 4) %>%
  ggplot(., aes(InquiriesLast6Months)) +
  geom_histogram(binwidth = 1) +
  theme_pander() +
  scale_x_continuous(breaks = seq(5, 27, 1))
```

The same data as table with the changed proportions is:

```{r table_inquiries_histogram_out_of_range}
post_sec_relaunch %>%
  filter(InquiriesLast6Months > 4) %>%
  count(InquiriesLast6Months) %>%
  mutate(freq = round(100 * (n / sum(n)), 2)) %>%
  kable(col.names = c('Inquiries Last 6 Months', 'Frequency', 'Proportion'))
```

These listings should be in status **cancelled** because they don't fullfill the
requirements to borrow money at Prosper.

```{r loan_status_out_of_range}
post_sec_relaunch %>%
  filter(InquiriesLast6Months > 4) %>%
  count(LoanStatus) %>%
  ggplot(., aes(LoanStatus, n)) +
  geom_bar(stat = 'identity') +
  theme_pander() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Obviously none of these listings are **cancelled**. It seems that this business
rule was not in place when the data was pulled or that all these loans are
older than 6 months.

## LoanOriginalAmount

The variable `LoanOriginalAmount` is of interest to see how much money is
borrowed. 

```{r loan_original_amount, results = 'asis'}
descr(post_sec_relaunch$LoanOriginalAmount, style = 'rmarkdown',
      omit.headings = TRUE)
```

The range of borrowed money is from **$1,000 to $35,000**. On average a listing 
is about **$9,092**. The median is at **$7,500**, the standard deviation is
**6288.78** and there are no **NA** values. 

The distribution is:

```{r loan_original_amount_hist}
# choose a histogram with binwidth eq 500 to see the peaks much better
ggplot(post_sec_relaunch, aes(LoanOriginalAmount)) +
  geom_histogram(binwidth = 500) +
  scale_x_continuous(label = dollar, breaks = seq(1000, 35000, 5000)) +
  theme_pander()
```

The plot is multimodal. It seems that most people choose standard amounts to
borrow at round numbers but that there are a some loans in between not using a
standard amount. Most loans are of **$4,000**, **$10,000** and **$15,000**.

How does that look like without these peaks?

```{r loan_original_amount_without_peaks}
post_sec_relaunch %>%
  count(LoanOriginalAmount) %>%
  filter(n < 500) %>%
  ggplot(., aes(LoanOriginalAmount)) +
  geom_histogram(binwidth = 500) +
  scale_x_continuous(label = dollar, breaks = seq(1000, 35000, 5000)) +
  theme_pander()
```

This plot is expected: It is right skewed. The higher the loan the fewer people
request that.

Let's squeeze together larger values and stretch out smaller values by using a
log scaling:

```{r log_loan_original_amount}
ggplot(post_sec_relaunch, aes(log(LoanOriginalAmount))) +
  geom_histogram(binwidth = 0.35) +
  theme_pander()
```

Using a log-Scale is multimodel as well but looks a bit more like a normal
distribution. This information might be handy if a model should be fit.

## LoanCurrentDaysDelinquent

`LoanCurrentDaysDelinquent` it the number of days the loan is delinquent.

```{r loan_current_days_delinquent_summary, results = 'asis'}
descr(post_sec_relaunch$LoanCurrentDaysDelinquent, style = 'rmarkdown', 
      omit.headings = TRUE)
```

The range of this variables is from **0** to **1593**. **1593** days are more
than **4 years**. On average a loan is **36.54** days deliquent. But as **Q1**,
Median and **Q3** are **0** the average is skewed from the high values. 
This variable contains no **NA** values. The distribution without 0-values is:

```{r loan_current_days_delinquent_distribution}
post_sec_relaunch %>%
  filter(LoanCurrentDaysDelinquent > 0) %>%
  ggplot(., aes(LoanCurrentDaysDelinquent)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0, 1600, 100)) +
  theme_pander()
```

As **0** is excluded the highest peaks are at the beginning of the plot and
around 115. The top 5 days delinquent are:

```{r loan_current_days_delinquent_peaks}
post_sec_relaunch %>%
  count(LoanCurrentDaysDelinquent) %>%
  top_n(., 5, n) %>%
  kable(col.names = c('Current Days Delinquent', 'Listings'))
```

This output shows the two peaks are at **10** and **121** days of delinquency.

```{r delayed_loans, results = 'hide'}
post_sec_relaunch %>%
  filter(LoanCurrentDaysDelinquent == 0) %>%
  nrow() / length(post_sec_relaunch$LoanCurrentDaysDelinquent)
```

As seen already in the summary statistics, most listings have a value of **0**
in `LoanCurrentDaysDelinquent`. Overall, roughly **90%** fall into this
category.

That means that about **10%** loans in this dataset are delinquent. According
to [Federal Reserve Economic Data][DRCLACBS] this is way more than compared to
all commercial banks in this timeframe of **3.19%**. That means that borrowers
at Prosper are more delinquent compared to commercial bank customers.

I'm left wondering how this information correlates to `BorrowerRate`, 
`ProsperScore` and `ProsperRating`. Do these 10% have a worse rating as well?

## CurrentDelinquencies

`CurrentDelinquencies` is the number of accounts delinquent at the time the
credit profile was pulled. What are the basic statistics?

```{r current_accounts_delinquent_summary, results = 'asis'}
descr(post_sec_relaunch$CurrentDelinquencies, style = 'rmarkdown', 
      omit.headings = TRUE)
```

As seen the range is **0** to **51**, **Q1** and **Q3** are **0** again. There
are no **NA** values, standard deviation is **1.11**.

I suspect that **51** accounts is a data issue. What is the listing having
this large number of delinquent accounts?

```{r maximum_current_delinquent}
post_sec_relaunch %>%
  filter(CurrentDelinquencies == max(CurrentDelinquencies))
```

From this data it does not look like an incorrect entry. `ListingCreationDate`
if from **2013-12-21**, `LoanStatus` is **current**, `ProsperScore` is **8**,
what is high (11 is the best). The borrower does not state his `Occupation` or
his `EmploymentStatusDuration` (what is set to **0**). He claims to be a
Homeowner with an `CreditScore` between **700** and **719**. His
`FirstRecordedCreditLine` is of **1979-07-19** what is **34 Years** ago when
this Prosper Credit was pulled. So the borrower is not young and it might be
that he really has an `IncomeRange` of **$50,000-74,999** due to his
`StatedMonthlyIncome` of **$5,000**.

But a little weird is that `TotalProsperLoans` and `TotalProsperPaymentsBilled`
is **NA** because `LoanMonthsSinceOrigination` is **3**. 

Currently it's unclear what this listing is about.

The distribution of this variable is:

```{r current_delinquencies_dist}
ggplot(post_sec_relaunch, aes(CurrentDelinquencies)) +
  geom_histogram(binwidth = 1) +
  theme_pander()
```

```{r current_delinquencies_zero, results = 'hide'}
post_sec_relaunch %>%
  filter(CurrentDelinquencies == 0) %>%
  nrow() / length(post_sec_relaunch$CurrentDelinquencies)
```

Again the majority of listings (**83%**) is having **0** current delinquencies. 

```{r current_delinquencies_dist_over_zero}
post_sec_relaunch %>%
  filter(CurrentDelinquencies > 0) %>%
  ggplot(., aes(CurrentDelinquencies)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(1, 51, 3)) +
  theme_pander()
```

Most of the members of these loans do have less than **10** delinquent accounts.

```{r more_than_10_delinquent_accounts}
post_sec_relaunch %>%
  filter(CurrentDelinquencies > 10) %>%
  ggplot(., aes(CurrentDelinquencies)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(11, 51, 3)) +
  scale_y_continuous(breaks = seq(0, 50, 2)) +
  theme_pander()
```

Plotting only the loans of members having more than **10** delinquent accounts
clearly show that the maximum is a large outlier. The next listings having a
large `CurrentDelinquencies` are **32**, **27** and **24**. That's alot.

## EstimatedLoss

Now let's see `EstimatedLoss`. This is defined as *the estimated principal loss
on charge-offs.* It is important because Prosper used a learning model on
historical data to calculate the estimated loss and uses this as one variable
to set its `ProsperScore`.

```{r estimated_los_summary, results = 'asis'}
descr(post_sec_relaunch$EstimatedLoss, style = 'rmarkdown', 
      omit.headings = TRUE)
```

As seen the range of this is from **0** to **0.37** or **37%**. The average
estimated loss is **0.08** or **8%**. 

```{r estimated_loss_distribution}
ggplot(post_sec_relaunch, aes(EstimatedLoss)) +
  geom_histogram(binwidth = 0.01) +
  theme_pander() +
  scale_x_continuous(label = percent)
```

The distribution shows that the majority of loans do have a estimated loss
below **20%**. Most loans do have a estimated loss of **6%** and there are some
loans above **20**. 

I wonder whether these can be related to other variables like `ProsperRating` or
`ProsperScore` as these are based on it.

## LoanStatus

The last variable I'd like to examine is `LoanStatus`. As this is a factor I
omit the descriptive statistics.

```{r loan_status}
ggplot(post_sec_relaunch, aes(LoanStatus)) +
  geom_bar() +
  geom_text(aes(y = ..count.., label = percent(..count.. / sum(..count..))), 
            stat = 'count', vjust = -0.5) +
  scale_y_continuous(breaks = seq(0, 60000, 10000)) +
  theme_pander() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

The majority of loans is **Current**, **23.1%** are **Completed** and about
**10%** are delayed. That was seen already above.

**1.2%** are **Defaulted**, **6.3%** are **Chargedoff**.

Later I'll correlating this with the other variables.

## Summary

In this first part 14 variables out of 81 in a large dataset of 113,937 loan
listings were examined. These 14 variables are:

|                      Variable |    Mean | Median |        Min |        Max | 
|------------------------------:|--------:|-------:|-----------:|-----------:|
|       **ListingCreationDate** |  &nbsp; | &nbsp; | 2009-07-29 | 2014-03-10 |
|              **BorrowerRate** |    0.20 |   0.19 |       0.04 |       0.36 |
|              **ProsperScore** |  &nbsp; |      6 |          1 |         11 |
|   **ProsperRating (numeric)** |  &nbsp; |  4 / C |     1 / HR |     7 / AA |
|     **CreditScoreRangeLower** |  &nbsp; |    700 |        600 |        880 |
|     **CreditScoreRangeUpper** |  &nbsp; |    719 |        619 |        899 |
|       **StatedMonthlyIncome** | 5932.96 |   5000 |          0 | 1750002.92 |
| **PublicRecordsLast12Months** |  &nbsp; |      0 |          0 |         20 |
|      **InquiriesLast6Months** |  &nbsp; |      0 |          0 |         27 |
|        **LoanOriginalAmount** | 9092.27 |   7500 |       1000 |      35000 |
| **LoanCurrentDaysDelinquent** |   36.54 |      0 |          0 |       1593 |
|      **CurrentDelinquencies** |  &nbsp; |      0 |          0 |         51 |
|             **EstimatedLoss** |    0.08 |   0.07 |          0 |       0.37 |
|                **LoanStatus** |  &nbsp; | &nbsp; |     &nbsp; |     &nbsp; |

`ProsperRating` is a numeric code as seen in the above table. The coding is

|  1 | 2 | 3 | 4 | 5 | 6 |  7 |
|---:|--:|--:|--:|--:|--:|---:|
| HR | E | D | C | B | A | AA |

* 1 is the worst
* 7 is the best rating

`ProsperScore` is the internal calculated score of a loan. It ranges from 1 to
11 whereby 11 is the best score.

Before doing any examinations the original dataset was reduced to include only
data after the relaunch due to the SEC registration. This was done to use only
the loans having a ProsperRating. This rating is used as primary risk analyzis
system after the SEC registration.

In addition to this manipulation `ProsperRating (Alpha)` was reordered from 
low to high risk (AA to HR or 7 to 1, respectively). This was done to display
this order more naturally.

For better displaying `LoanStatus` this was reorderd from *god* to *bad* status
as well. That means from `Completed` to `Chargedoff`. 

During the analyzis my observations were:

* Most loans have a interest rate of **32%**.
* About 58% of the loans of the top 5 highest interest rates do have the worst
  rating of **HR** or *High Risk*.
* The majority of loans do have a `ProsperScore` of **4**, **6** and **8**. 
* The modal of `ProsperRating` is **4** - meaning that most loans do have a
  **C** rating. This variable is the only one that is normal distributed.
* Prosper does not allow members having a `CreditScore` below **600**. Most
  members do have a score of **660-679**.
* The average stated monthly income is **$5,932.96** but there are **21048**
  listings having stated more than **$7,083.33**. As seen above there is one
  loan having a stated income of **$1,750,002**. I don't think that someone
  having that large income per month would borrow money at Prosper. So I
  suspect that a outlier. But this can not be proven.
* There are **682** listings having at least one public record within the last
  12 months. This is prohibited by Prosper at the time joining the Marketplace.
  If necessary for further analyzis it should be checked whether these loans
  are older than 12 months. That would mean that the public records were filed
  during the term of the loan.
* **97%** of the loans did not have any inquiries within the last 6 months.
* Most borrowers borrowed **$4,000**, **$10,000** or **$15,000**. Very few
  listings exist having borrowed more than **$25,000**.
* The average estimated loss is **8%**. 
* Most loans do have an estimated loss below **20%**, the most of **6%**.
* About **90%** of the loans are not delinquent. That means that **10%** are
  delinquent. That is more than the **3.19%** average borrowers using
  commercial banks.

The main features in this dataset are `BorrowerRate` and `ProsperRating`. I 
suspect a high correlation between them. 

I suspect all other variables expect `PublicRecordsLast12Months` to be 
helpful analyzing `BorrowerRate` and `ProsperRating`. I think that all of these
contribute to them.

```{r cleanup_part_1}
rm(rate_order, top_5_worse_borrower_ratings)
```

***

# Let's go swimming

The second part of this report examines correlation between above variables to
find patterns and perhaps get a glimpse of predicting risk factors.

## Correlogram

Let's start with a correlogram

```{r correlogram}
post_sec_relaunch %>%
  select(BorrowerRate, ProsperScore, ProsperRating..numeric.,
         CreditScoreRangeLower, StatedMonthlyIncome, PublicRecordsLast12Months,
         InquiriesLast6Months, LoanOriginalAmount, LoanCurrentDaysDelinquent,
         CurrentDelinquencies, EstimatedLoss) %>%
  cor() %>%
  ggcorrplot(., type = 'full', ggtheme = theme_pander(), hc.order = TRUE,
             title = 'Correlogram of Loan Data', lab = TRUE, lab_size = 2.5,
             colors = c("#e03c31", "white", "#a4c639"), tl.cex = 10, 
             show.diag = TRUE)
```

Based on this correlogram the strongest relationships of `BorrowerRate` are

* `ProsperRating`
* `EstimatedLoss`

`ProsperRating` has the strongest relationships with

* `EstimatedLoss`
* `BorrowerRate`
* `ProsperScore`

`ProsperScore` has relationships with

* `EstimatedLoss`
* `BorrowerRate`
* `ProsperRating`

Let's explore some of these relationships. As the primary variable of interest
is `BorrowerRate` I start with it.

## BorrowerRate and ProsperRating

Based on the correlogram this is a strong relation in the examined variables.

```{r borrower_rate_by_prosper_rating}
ggplot(post_sec_relaunch, aes(ProsperRating..numeric., BorrowerRate)) +
  geom_jitter(alpha = 0.25, size = 0.5) +
  theme_pander() +
  scale_y_continuous(label = percent) +
  scale_x_continuous(breaks = seq(1, 7, 1))
```

As this plot already looks like a boxplot let's choose `ProsperRating (Alpha)`
in a second plot. 

```{r borrower_rate_by_prosper_rating_alpha}
ggplot(post_sec_relaunch, aes(ProsperRating..Alpha., BorrowerRate)) +
  geom_boxplot() +
  theme_pander() +
  scale_y_continuous(label = percent) +
  # as ProsperRating (Alpha) was ordered from Low to High Risk it must be
  # reversed here
  scale_x_discrete(limits = rev(levels(post_sec_relaunch$ProsperRating..Alpha.)))
```

As expected the correlation can be seen here. The better `ProsperRating` the 
smaller `BorrowerRate`. Except for **HR** all *IQR*s even don't overlap.

Let's the what correlates to `ProsperRating`.

## ProsperRating and EstimatedLoss

`ProsperRating` has a strong relation to `EstimatedLoss`.

```{r prosper_rating_by_estimated_loss}
ggplot(post_sec_relaunch, aes(ProsperRating..Alpha., EstimatedLoss)) +
  geom_tufteboxplot() +
  scale_y_continuous(label = percent) +
  theme_pander() +
  scale_x_discrete(limits = rev(levels(post_sec_relaunch$ProsperRating..Alpha.)))
```

This can be seen by this plot. The better `ProsperRating` the lower
`EstimatedLoss`.

So what about the correlations of `EstimatedLoss`?

## EstimatedLoss and ProsperScore

According to [Prosper][GPS] `EstimatedLoss` was used to build their custom risk
score.

```{r estimated_loss_by_prosper_score}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), EstimatedLoss)) +
  geom_tufteboxplot() +
  scale_y_continuous(label = percent) +
  theme_pander()
```

And yes: This is visible here. The higher `ProsperScore` the less is
`EstimatedLoss`. 

## EstimatedLoss and CreditScore

One variable related to `EstimatedLoss` should be `CreditScoreRangeLower` and
`CreditScoreRangeUpper` because that is a hint of creditworthiness.

```{r estimated_loss_by_credit_score}
ggplot(post_sec_relaunch, aes(CreditScoreRangeLower, EstimatedLoss)) +
  geom_jitter(alpha = 0.25, size = 0.5) +
  geom_smooth(color = 'red', size = 0.75, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_y_continuous(label = percent) +
  theme_pander()
```

That's the expected correlation. The higher the credit score the lower the 
estimated loss. 

In my opinion the correlations are: If `ProsperRating` is low the loan gets a 
better `BorrowerRate`. `ProsperRating` is lower if `EstimatedLoss` is low. And
`EstimatedLoss` is low if (a) the `CreditScore` is high and (b) the custom risk
score `ProsperScore` is high (11 being the best).

I suspect these variables to have a relationship to `ProsperScore`:

* `InquiriesLast6Months` is stated by Prosper to be a prerequisite to join the
  Marketplace at all.
* `CurrentDelinquencies` because this might be a hint on the creditworthiness.
* `AvailableBankcardCredit` and `BankcardUtilization` - not examined so far - 
  might be a hint about the creditworthiness as well.
* `TotalProsperLoans` because I suspect that Prosper takes it into account
  whether it knows the member already.

## ProsperScore and InquiriesLast6Months

As written Prosper takes into account the number of inquiries on the credit 
bureau. Because of that let's see the relationship between 
`InquiriesLast6Months` and `ProsperScore`.

```{r prosper_score_by_inquiries_last_6_months}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), InquiriesLast6Months)) +
  geom_boxplot() +
  theme_pander()
```

Again the plot is skewed to outliers or higher edge cases. 

```{r prosper_score_by_inquiries_last_6_months_subset}
# I choose a jitter plot here because boxes do not show the trend that much
post_sec_relaunch %>%
  filter(InquiriesLast6Months < 12) %>%
  ggplot(., aes(as.factor(ProsperScore), InquiriesLast6Months)) +
  geom_jitter(size = 0.5) +
  scale_y_continuous(breaks = seq(0, 12, 1)) +
  theme_pander()
```

Using a subset of data shows the trend more clear. The more inquiries exist the
lower `ProsperScore`.

## ProsperScore and CurrentDelinquencies

As every Bank does pay attention to the number of delinquent accounts of its
customers I suspect that Prosper is doing that as well. Next I plot
`CurrentDelinquencies` and `ProsperScore`.

```{r prosper_score_by_current_delinquencies}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), CurrentDelinquencies)) +
  geom_boxplot() +
  theme_pander()
```

This plot looks normal distributed. Let's remove the higher values.

```{r prosper_score_by_current_delinquencies_subset}
# again I choose a jitter plot because a boxplot or violin plot is not really 
# helpfull in plotting a trend here. That' because most values are 0
post_sec_relaunch %>%
  filter(CurrentDelinquencies < 20 &
         CurrentDelinquencies > 0) %>%
  ggplot(., aes(as.factor(ProsperScore), CurrentDelinquencies)) +
  geom_jitter(alpha = 0.25, size = 0.5) +
  theme_pander()
```

Again this Plot looks normal distributed. So I suspect this variable is not
that correlated to `ProsperScore`.

## ProsperScore and AvailableBankcardCredit

I suspect that correlates as well strong. Having a higher credit available is a
hint of a better solvency.

```{r prosper_score_by_available_card_credit}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), AvailableBankcardCredit)) +
  geom_boxplot() +
  scale_y_continuous(label = dollar) +
  theme_pander()
```

As there are high values let's remove them.

```{r prosper_score_by_available_card_credit_subset}
post_sec_relaunch %>%
  filter(AvailableBankcardCredit < 100000) %>%
  ggplot(., aes(as.factor(ProsperScore), AvailableBankcardCredit)) +
  geom_boxplot() +
  scale_y_continuous(label = dollar) +
  theme_pander()
```

Again the trend seems to confirm my assumption. The higher the available
bankcard credit the higher the `ProsperScore`.

## ProsperScore and BankcardUtilization

I suspect a better solvency if one did not use his bankcard that often.

```{r prosper_score_by_bankcard_utilization}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), BankcardUtilization)) +
  geom_boxplot() +
  theme_pander()
```

I don't think there are much outliers. But I'll filter the data nevertheless.

```{r prosper_score_by_bankcard_utilization_subset}
post_sec_relaunch %>%
  filter(BankcardUtilization < 1.25) %>%
  ggplot(., aes(as.factor(ProsperScore), BankcardUtilization)) +
  geom_boxplot() +
  theme_pander()
```

A trend can be seen that a smaller `BankcardUtilization` leads to a higher
and hence better `ProspeScore`.

## ProsperScore and TotalProsperLoans

Let's see the correlation of these two variables. I think that Prosper 
calculates `ProsperScore` different if a loan listing is files by an existing
member.

```{r prosper_score_by_total_prosper_loans}
ggplot(post_sec_relaunch, aes(as.factor(ProsperScore), TotalProsperLoans)) +
  geom_violin(na.rm = TRUE, trim = FALSE, scale = 'count') +
  theme_pander()
```

Again I don't see correlation here. I assume that this follows a normal
distribution.

## Relations of ListingCreationDate

Just of interest the last analyzis in part two is about the trend of 
`BorrowerRate`, `ProsperRating` and `LoanCurrentDaysDelinquent` over
time.

### BorrowerRate

```{r borrower_rate_by_day}
post_sec_relaunch %>%
  group_by(ListingCreationDate) %>%
  summarise(avgBorrowerRate = mean(BorrowerRate)) %>%
  select(ListingCreationDate, avgBorrowerRate) %>%
  ggplot(., aes(ListingCreationDate, avgBorrowerRate)) +
  geom_line() +
  geom_smooth(color = 'red', size = 0.5, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_y_continuous(label = percent) +
  theme_pander()
```

Before 2011 there was a little noise in the mean borrower rate. In 2011 and 2012
the plot looks a little seasonal. But after the first half of 2011 the trend is
downward. Can that be related by the `ProsperRating`?

### ProsperRating

Let's plot the same with `ProsperRating`.

```{r prosper_rating_by_day}
post_sec_relaunch %>%
  group_by(ListingCreationDate) %>%
  summarise(avgProsperRating = mean(ProsperRating..numeric.)) %>%
  select(ListingCreationDate, avgProsperRating) %>%
  ggplot(., aes(ListingCreationDate, avgProsperRating)) +
  geom_line() +
  geom_smooth(color = 'red', size = 0.5, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  theme_pander()
```

My assumption seems to be confirmed by this plot: Starting in 2011 the trend of
the average ProsperRating is upwards.

### LoanCurrentDaysDelinquent

The next plot in this section is about the trend in delinquent loans:

```{r delinquent_loans_by_day}
post_sec_relaunch %>%
  mutate(BadLoan = 
           if_else(.$LoanStatus %in% c('Completed', 'FinalPaymentInProgress', 
                                       'Current'), FALSE, TRUE, TRUE)) %>%
  group_by(ListingCreationDate) %>%
  summarise(avgBadLoans = mean(BadLoan)) %>%
  select(ListingCreationDate, avgBadLoans) %>%
  ggplot(., aes(ListingCreationDate, avgBadLoans)) +
  geom_line() +
  geom_smooth(color = 'red', size = 0.5, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_y_continuous(label = percent) +
  theme_pander()
```

This trend is obvious: Since the second half of 2011 the average number of
delinquent loans per day is downwards. 

It seems that the risk evaluating system of Prosper is doing well.

## Summary

In the second part I first draw a correlogram of all variables choosen in the
first part of the document. For my luck these were only numeric variables
containing no *NA* values. 

The primary variable examined was `BorrowerRate`. The first correlation
examinded was with `ProsperRating` because this is the strongest having a 
correlation coefficient of **-0.95**. This negative correlation was seen in the
plot. 

The question I wanted to answer is: What contributes to a low `BorrowerRate`.
Hence I followed the track to plot the highest correlation of `ProsperRating`.
This is `EstimatedLoss`. This plot shows that a better `ProsperRating` 
correlates with a lower `EstimatedLoss`. 

Again I took the strongest correlation of `EstimatedLoss`. This is 
`ProsperScore`. This plot shows that a higher `ProsperScore` contributes 
to a lesser `EstimatedLoss`.

According to Prosper the second variable that contributes to a lesser 
`EstimatedLoss` is the `CreditScore`. Hence I plottet this relationship. I found
the expected correlation: The higher the credit score the lower the estimated 
loss.

The next stept was to identify variables having a relationship to 
`ProsperScore` because this the most important variable to set `EstimatedLoss`
and hence `ProsperRating`. I examined `InquiriesLast6Months`, 
`CurrentDelinquencies`, `AvailableBankcardCredit`, `BankcardUtilization` and
`TotalProsperLoans`.

The obvservations made are:

* The higher `InquiriesLast6Months` the lower is `ProsperScore` and hence I
  assume the loan is rated worst.
* There is no significant correlation with `CurrentDelinquencies`.
* The higher the `AvailableBankcardCredit` the higher the `ProsperScore`.
* A smaller `BankcardUtilization` leads to a higher and hence better 
  `ProspeScore`.
* I don't see a significant correlation with `TotalProsperLoans`. 

In my opinion a lower `BorrowerRate` is offered to borrowers that

* have no or only a few inquiries within the last 6 month, 
* have available a high bankcard credit,
* have a low bankcard utilization,
* have a high credit score.

In addition three plots were created to show the trend of `BorrowerRate`,
`ProsperRating` and `LoanCurrentDaysDelinquent`.

* The average `BorrowerRate` dropped from the second half of 2011.
* Starting in 2011 the trend of the average ProsperRating is upwards.
* Since the second half of 2011 the average number of delinquent loans per day 
  is downwards.

# Scuba Diving

In this part of the document I'll try to prove the findings of the sections
above. As it turned out that the primary variables are `BorrowerRate` and 
`EstimatedLoss`. I'm going to examine these further.

## EstimatedLoss, ProsperScore and CreditScore

`EstimatedLoss` is based on `ProsperScore` and `CreditScore`. So let's first
plot these.

```{r estimated_loss_by_credit_score_and_prosper_score}
ggplot(post_sec_relaunch, aes(EstimatedLoss, CreditScoreRangeLower,
                              colour = as.factor(ProsperScore))) +
  geom_jitter(alpha = 0.25, size = 0.5) +
  geom_smooth(color = 'red', size = 0.75, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_x_continuous(label = percent) +
  scale_colour_brewer(type = 'div') +
  theme_pander()
```

This relationship can be seen in this plot. The lower the credit score the
higher `EstimatedLoss`. In addition it can be seen that the higher prosper
scores are plotted at the top left edge with the lower estimated loss.

## EstimatedLoss, InquiriesLast6Months and AvailableBankcardCredit

As `EstimatedLoss` is based on `ProsperScore` as well as `CreditScore` and 
`ProsperScore` in turn correlates to `InquiriesLast6Months`, 
`AvailableBankcardCredit` and `BankcardUtilization` this should be visible as
well without using `ProsperScore`.

```{r estimated_loss_and_il6m_and_abc}
ggplot(post_sec_relaunch, aes(EstimatedLoss, InquiriesLast6Months,
                              colour = AvailableBankcardCredit)) +
  geom_point(size = 0.5) +
  geom_encircle(data = subset(post_sec_relaunch, InquiriesLast6Months > 15),
                color = '#0070ff', expand = 0.05) +
  # choose the same subset as above: $150,000
  geom_encircle(data = subset(post_sec_relaunch, 
                              AvailableBankcardCredit > 150000),
                color = '#4b5320', expand = 0.05) +
  scale_x_continuous(label = percent) +
  scale_y_continuous(breaks = seq(0, 35, 5)) +
  scale_color_gradient(low = '#f5f5dc', high = '#c32148') +
  theme_pander()
```

In this plot all listings having more than **15** inquiries are encircled in
blue. All listings having more than **$150,000** available bankcard credit are
encircled green and all listings having a high amount of available bankcard
credit are filled red. These are visible in the lower left of the plot - but as
there are listings with a very high amount of credit only a few points are
filled dark red.

As visible the correlation between these three variables exist the same way as
between `ProsperScore`, `InquiriesLast6Months` and `AvailableBankcardCredit`.

## BorrowerRate, EstimatedLoss and InquiriesLast6Months

As seen the correlation is even visible by excluding `ProsperScore`. I assume
that this is also visible by using `BorrowerRate` directly. 

To recap first: 

* `BorrowerRate` correlates with `ProsperRating`. 
* `ProsperRating` correlates with `EstimatedLoss`.
* `EstimatedLoss` correlates with `ProsperScore`. 
* `ProsperScore` correlates among others with `InquiriesLast6Months`.

A correlation should be visible between `BorrowerRate`, `EstimatedLoss` and
`InquiriesLast6Months`.

To reduce overplotting I exclude all loans having less than 6 inquiries in the
last 6 months.

```{r borrower_rate_plus_estimated_loss_plus_inquiries_last_6_months_subset}
post_sec_relaunch %>%
  filter(InquiriesLast6Months > 5) %>%
  ggplot(., aes(BorrowerRate, EstimatedLoss, color = InquiriesLast6Months)) +
  geom_point(aes(size = InquiriesLast6Months)) +
  geom_smooth(color = 'red', size = 0.75, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_y_continuous(label = percent) +
  scale_x_continuous(label = percent) +
  theme_pander() +
  scale_color_gradient(low = '#f5f5dc', high = '#c32148')
```

It is visible that the higher the borrower rate and the higher the estimated
loss the bigger the points. Hence the more inquiries were done. 

My assumptions seem to be correct.

## BorrowerRate, EstimatedLoss and ProsperRating

As last plot let's include `ProsperRating` instead of `InquiriesLast6Months`.
Doing that shows the two strongest correlations.

```{r borrower_rate_plus_estimated_loss_plus_prosper_rating}
ggplot(post_sec_relaunch, aes(BorrowerRate, EstimatedLoss)) +
  geom_density2d(aes(colour = ProsperRating..Alpha.)) +
  scale_x_continuous(label = percent) +
  scale_y_continuous(label = percent) +
  theme_pander() +
  scale_colour_brewer(type = 'qual', palette = 2)
```

This plot shows very clear that listings having a low `EstimatedLoss` and
a high `ProsperRating` do have a low `BorrowerRate`.

## Summary

In this third part of the document I verified the correlations found in the
second part. 

First I checked correlation between `EstimatedLoss`, `ProsperScore` and 
`CreditScore` because the former variable is calculated by the other two. This
can be seen in the plot. The higher `CreditScore` and the higher `ProsperScore`
the lower `EstimatedLoss`. 

After that I checked whether the correlations of `ProsperScore` with
`InquiriesLast6Months` and `AvailableBankcardCredit` can be seen as well for
`EstimatedLoss`. I expected that due to the correlations mentioned above. This
was confirmed by the corresponding plot.

The third correlation checked was between `BorrowerRate`, `EstimatedLoss` and 
`InquiriesLast6Months` because `BorrowerRate` highly correlates with 
`EstimatedLoss` and `EstimatedLoss` highly correlates with 
`InquiriesLast6Months`. These correlations were again confirmed. 

The last plot related `BorrowerRate`, `EstimatedLoss` and `ProsperRating` to
validate whether these correlations are strong.

I confirmed that `BorrowerRate` depends on `InquiriesLast6Months`,
`AvailableBankcardCredit` and `BankcardUtilization`.

These are for sure not the only variables used to calculate `BorrowerRate` but
they do have a strong relationship.

***

# Final Plots and Summary

## Average Number of Loans delayed per Day

```{r final_delinquent_loans_by_day}
post_sec_relaunch %>%
  mutate(BadLoan = 
           if_else(.$LoanStatus %in% c('Completed', 'FinalPaymentInProgress', 
                                       'Current'), FALSE, TRUE, TRUE)) %>%
  group_by(ListingCreationDate) %>%
  summarise(avgBadLoans = mean(BadLoan)) %>%
  select(ListingCreationDate, avgBadLoans) %>%
  ggplot(., aes(ListingCreationDate, avgBadLoans)) +
  geom_line(color = '#120a8f') +
  geom_smooth(color = '#f3e5ab', size = 0.75, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_y_continuous(label = percent) +
  labs(x = 'Listing Date', y = 'Average Number of Loans delayed', 
       title = 'Average Number of Loans delayed per Day',
       subtitle = 'Listings from 2009-07-29 to 2014-03-10') +
  theme_pander()
```

## Description

This plot shows the average number of Loans per Day that are delayed. That means
Loans that are past due, defaulted or chargedoff.

One can see here that Prosper implemented a risk estimation system
that started to work better and better starting from half of 2011. Prosper used
the data of their population from April, 2008 through April, 2011 to build this
system and enhanced it over time.

## Estimated loss by Credit Score and Prosper Score

```{r final_estimated_loss_by_credit_score_and_prosper_score}
ggplot(post_sec_relaunch, aes( EstimatedLoss, CreditScoreRangeLower,
                              colour = as.factor(ProsperScore))) +
  geom_jitter(alpha = 0.25, size = 0.5) +
  geom_smooth(color = 'red', size = 0.75, method = 'gam',
              formula = y ~ s(x, bs = 'cs')) +
  scale_x_continuous(label = percent) +
  labs(title = 'Estimated loss by Credit Score and Prosper Score', 
       x = 'Estimated Loss', y = 'Credit Score', 
       colour = 'Prosper Score\nBest = 11') +
  scale_colour_brewer(type = 'div') +
  theme_pander()
```

## Description

The risk estimation system of Prosper is based on the calculation of the
variable `EstimatedLoss`. These is in turn based on the two variables
`ProsperScore` and `CreditScore`.

This can be seen in this plot. The lower the credit score and Prosper Score the
higher `EstimatedLoss`. And the higher `EstimatedLoss` the higher 
`ProsperRating` and hence `BorrowerRate`. This can be seen in the next plot.

## Borrower Rate by Estimated Loss and Prosper Rating

```{r final_borrower_rate_plus_estimated_loss_plus_prosper_rating}
ggplot(post_sec_relaunch, aes(BorrowerRate, EstimatedLoss)) +
  geom_density2d(aes(colour = ProsperRating..Alpha.)) +
  scale_x_continuous(label = percent) +
  scale_y_continuous(label = percent) +
  labs(title = 'Borrower Rate by Estimated Loss and Prosper Rating', 
       x = 'Borrower Rate (interest rate)', y = 'Estimated Loss', 
       colour = 'Prosper Rating') +
  theme_pander() +
  scale_colour_brewer(type = 'qual', palette = 2)
```

## Description

As described above the interest rate that Prosper offers its members is based
on the calculation of the loss if a loan is chargedoff. This loss is condensed
in a rating from **AA** (lowest risk) to **HR** (highest risk). And based on
this rating the borrower rate is set.

This plot shows this relationship. One can see that the lower borrower rates are
set on `ProsperRating` **AA** which has the lowest `EstimatedLoss`.

***

# Reflection

Analyzing this dataset was a big challenge. Not due to the size of data. These
large number of observations can be handled with the computing power at hand.
Even the number of variables was not a problem concerning computations. 

The biggest issue was the missing domain knowlegde. I did not know Prosper and
don't know that much about finance. And this was a problem having 81
variables available. In addition I'm not a native speaker of English. So I
did not know what 'interest rate' is or the meaning of 
'Total principal borrowed on Prosper loans'.

During the project I've thrown away three different versions of the report just
because I learned more and more about the structure and the relevance of the
variables. And because I saw the mistakes I made due to incorrect presumptions
and translations.

Another struggle was the syntax and funtionality of R. Subsetting and plotting
data is easier in Python in my opinion. At the beginning of the project a lot
of time was spent to forget what I've learned in Term 1 about Python and
learn how I can achieve the same with R. 

Finally I think I was able to gain a little insight into the business of
P2P lending and how Prosper used existing data to predict the loan outcome of
its members.

In the future I'd try to find more variables contributing to `EstimatedLoss` and
fit a model with a larger dataset. In addition I'm left wondering what the 
perfect portfolio of Prosper loans look like to maximize the lenders yield.

***

# References

* [Investopia](https://www.investopedia.com)
* [Prosper Marketplace at WikiPedia](https://en.wikipedia.org/wiki/Prosper_Marketplace)
* [Harvard Business Law Review](http://www.hblr.org/2016/08/it-aint-broke-the-case-for-continued-sec-regulation-of-p2p-lending/)
* [Prosper Minimum Borrowing Criteria](https://prosper.zendesk.com/hc/en-us/articles/210013963-What-are-the-minimum-criteria-to-borrow-on-Prosper-)
* [dplyr Reference](https://dplyr.tidyverse.org/reference/index.html)
* [ggplot2 Referece](https://ggplot2.tidyverse.org/reference/index.html)
* [Prosper Score](https://www.prosper.com/plp/general-prosper_score/)
* [Estimated Loss Rates](https://www.prosper.com/plp/general-estimated_loss_rates/)
* [Prosper Ratings](https://www.prosper.com/plp/invest/prosper-ratings/)

[WikiPedia]: https://en.wikipedia.org/wiki/Prosper_Marketplace
[HBLR]: http://www.hblr.org/2016/08/it-aint-broke-the-case-for-continued-sec-regulation-of-p2p-lending/
[MinimumCriteria]: https://prosper.zendesk.com/hc/en-us/articles/210013963-What-are-the-minimum-criteria-to-borrow-on-Prosper-
[DRCLACBS]: https://fred.stlouisfed.org/series/DRCLACBS
[TERMCBCCALLNS]: https://fred.stlouisfed.org/series/TERMCBCCALLNS
[GPS]: https://www.prosper.com/plp/general-prosper_score/
